---
title: "Analysis of Variance"
subtitle: "In Class Activity"
author: "Emma Sharp"
date: "11/20/2024"
execute: 
  echo: true
format: 
  html:
    code-tools: true
    toc: false
editor_options: 
  chunk_output_type: console
---

Here is some new data to play with a bit to explore one-way and two-way analysis of variance models.  These data represent weight data associated with three different diet types.  There are 76 observations of individuals with the following characteristics:

- Gender (0/1)
- Age (earth years. 🤓)
- Height (cm)
- Pre.weight & weight6weeks (kg)
- Diet (1,2,3)

The data are located as a CSV file named `DietData.csv` linked from the Canvas page.  Load the data in and format things so they look correctly to you.

```{r}
#| message: false
# load the data in here
library(readr)
library(tidyverse)
library(knitr)

diet <- read_csv("DietData.csv") %>% 
  mutate(weight_loss = pre.weight - weight6weeks,
         Diet = as.factor(Diet),
         Gender = ifelse(gender == "0",
                         "Female",
                         "Male"))
```

## Exploratory Data Anlsysis

One of the first things to do is to look at the data and see if there are any obvious things.  Go ahead and explore these data visually.  What do you see?

```{r}
# Data visualization

hist(diet$weight_loss)

diet %>% 
  ggplot(aes(x = Diet,
             y = weight_loss))+
  geom_boxplot()+
  labs(title = "Diet 3 Seems to be Most Effective, with Diet 1 & 2 Being Pretty Similar")

diet %>% 
  ggplot(aes(x = Gender,
             y = weight_loss))+
  geom_boxplot()+
  labs(title = "Weight Loss Across Gender Seems to be Pretty Similar")

diet %>% 
  ggplot(aes(x = Gender,
             y = pre.weight))+
  geom_boxplot()+
  labs(title = "Starting Weight is Much Higher for Men than it is for Women")
```

## Estimating Mean Values

Make a table of Weight Loss by gender and diet.

```{r}
#| warning: false
# Table output

diet %>% 
  group_by(Diet, Gender) %>% 
  summarise(`Weight Loss` = mean(weight_loss)) %>% 
  kable()
```


## 1-Way Analysis of Variance

*Diet Issues:* 


The underlying linear model.


$$
y_{ij} = \mu + \tau_{Diet, i} + \epsilon_j
$$


Test the null hypothesis, $H_O:$ *There is no effect of diet on weight loss* (e.g., $\tau_{Diet-1} = \tau_{Diet-2} = \tau_{Diet-3} = 0.0$).  Is there evidence for one diet producing more weight loss than the others?  Create an `aov` analysis and assign it to the variable `fit.diet` and examine its contents.

```{r}
# Define model

fit.diet <- aov(weight_loss ~ Diet, data = diet)
anova(fit.diet)

# Because the p value is 0.006, we can reject the null hypothesis.  This means that at least one of the diets has a significantly different mean weight loss than the others. 
```

Are they all significantly different?  Try the `TukeyHSD()` Interpret the results.

```{r}
# Posthoc test

TukeyHSD(fit.diet)

#Diet 1 and diet 2 do not have significantly different means.  However, diet 3 has a significantly higher mean than both diet 1 and diet 2.
```

How much of the variation is explained?  If you notice when you do a summary from a `lm()` (regression) model, it gives you the $R^2$ values directly (remember $R^2 = \frac{SS_{model}}{SS_{Error}}$).  Does `summary()` of your `aov` model give you that?
 
```{r}
# How much variance?

summary(fit.diet)

#The summary of the aov doesn't give us a r^2 value.
```

Since I asked the question, the answer is probably no.  Why does it not do this?  Probably for historical reasons, which are a bit of a pain in the backside.  That being said, there are some `tidy` ways to fix this issue.  I'm going to use the `broom` package which allows us to clean up (or tidy if you will) the model objects.  This will take the model object and pull out *all* the ANOVA table stuff and put it into a `tibble`.

```{r}
library( broom ) 
# use your model fit next (I called mine fit.diet)
tidy_diet <- tidy( fit.diet )
tidy_diet
```

Now, since it is all easily accessible, we can calculate the $R^2$ from the new model output.

```{r}
# Estimate the variance explained from the raw sums of squares

r2_Diet <- tidy_diet$sumsq[1] / sum( tidy_diet$sumsq )
r2_Diet 
```


 

*Gender:* 

The underlying linear model.


$y_{ij} = \mu + \tau_{gender, i} + \epsilon_j$



Independent of the diet, test the null hypothesis $H_O:$ *There is no difference in weight loss between genders* (e.g., $\tau_{gender-0} = \tau_{gender-2} = 0.0 $).  Is there evidence for one gender being significantly different than another?  How much of the variation is explained (another $R^2$ by gender)?

```{r}
# partition effects

fit.gender <- aov(weight_loss ~ Gender, data = diet)
anova(fit.gender)

#There is not enough evidence (p = 0.83) to reject the null hypothesis that there is no difference in weight loss between genders.
```

How do you interpret these results thus far?

Diet 3 seems to be the most effective diet for weight loss.  Diet 1 and 2 are both similar in terms of weight loss.  Gender does not seem to play a role in how much weight a person is going to lose.

# Do genders respond differently to diets?


$y_{ijk} = \mu + \tau_{Diet,i} + \tau_{gender,j} + \epsilon_k$

So here we do something a bit different.  We want to simultaneously ask the following questions:

- Do diets influence weight loss?
- Do genders influence weight loss?
- Is there an interaction where different genders respond differently to different diets?

In $R$, this is done as:

```{r}
# diet model
anova(fit.diet)
```

```{r}
# gender model
anova(fit.gender)
```

```{r}
# interaction model
fit.both <- aov(weight_loss ~ Diet + Gender + Diet * Gender,
                data = diet)
anova(fit.both)

tidy_both <- tidy(fit.both)
tidy_both

r2_both <- tidy_both$sumsq[1] / sum(tidy_both$sumsq)
r2_both
```

What is the $R^2$ for this model?

0.129

```{r}
# best model variance explained

data.frame(model = c("Diet", "Gender", "Both"), 
           AIC = c(AIC(fit.diet), AIC(fit.gender), AIC(fit.both)))

```


# Which Model is Best?

How would you compare the models you generated?  How do you interpret the findings?

*Based on the AIC, there isn't one best model, but two.  The two best models are the relationship between diet type and weight loss as well as the relationship between the interactions between diet type and gender and weight loss.  The model only based on Diet type has a p value of 0.007, so we can conclude that at least one of the diets produces a different mean weight loss value.  Because of the post-hoc test, we know that diet 3 produces a significantly higher weight loss than the other two diets.  The interaction model produces significant p values for the diet term alone (p = 0.005) and for the interaction term (p = 0.05), but not for gender alone (p = 0.86).*

