---
title: "Correlations"
author: "Emma Sharp"
format: html
editor: visual
date: "11/08/2024"
editor_options: 
  chunk_output_type: console
---
## Correlation Assignment:

```{r}
#| include: false
source("get_rice_data.R")
rice <- get_rice_data()
library(GGally)
library(tidyverse)
```

1. Look up the library GGally; it has a function named ggpairs(). Use that function to plot the atmospheric data from the Rice Rivers Center for the samples collected from 10 February to 15 February.  

```{r}
rice %>% 
  filter(Date >= "2014-02-10 00:00:00" &
           Date <= "2014-02-15 24:60:60") -> rice2
  
rice2 %>% 
  select(AirTempC, WindSpeed_mph, Rain_cm, RelHumidity, BP_HG) %>% 
  ggpairs()
```

2. For those atmospheric data, which pair of variables has the strongest correlation? What is the 95% confidence interval on that correlation coefficient?

Barometric pressure and air temperature had the strongest correlation at -0.692.  The 95% confidence interval on that correlation coefficient is -0.736 to -0.642.

```{r}
cor.test(rice2$AirTempC, rice2$BP_HG)
```


3. Using the first 40 observations in air temperature and barometric pressure from the Rice Center data set, determine if they are individually distributed as normal random variables.  

```{r}
heads <- head(rice, 40) 

hist(heads$AirTempC)
shapiro.test(heads$AirTempC) #air temp is normally distributed

hist(heads$BP_HG)
shapiro.test(heads$BP_HG) #barometric pressure isn't normally distributed 
```

4. Given your findings in the last question, what kind of correlation statistic would be most appropriate for estimating the correlation between this subset of data?

Because the barometric pressure isn't normally distributed, the correlation statistic most appropriate would be the Spearman's Rho test.  You could also use a permutation.

5. Look at a qqnorm() plot of the barometric pressure data you used in the previous example. Is there something that "looks" odd with these data? Explain why those data are the way they are. 

It looks like the data trails off at the edge of the graph instead of staying close to the line.  This could be because of the way the data is distributed.

```{r}
qqnorm(heads$BP_HG)
qqline(heads$BP_HG, col = "red")
```

6. Using a permutation approach, define the distribution of correlation values between the variables in #3, assuming that the NULL hypothesis is true. Plot these as a histogram and include the observed correlation.

```{r}
df <- data.frame(Estimate = factor(c("Original",
                                     rep("Permuted", 999))), 
                 rho = c(cor.test(heads$AirTempC, heads$BP_HG)$estimate,
                         rep(NA, 999)))

for( i in 2:1000) {
  yhat <- sample(heads$BP_HG,
                 size = length(heads$BP_HG),
                 replace = FALSE)
  model <- cor.test(heads$AirTempC, yhat)
  df$rho[i] <- model$estimate 
}

ggplot(df)+
  geom_histogram(aes(rho,
                     fill = Estimate))
```

Because the original is outside of the 95% percentile of the permutated values, there is enough evidence to reject the null hypothesis.