---
title: "Rasters"
author: "Emma Sharp"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

1. Load the raster and point data in and crop to an appropriate size to display the locations of the sampling plots and make a label for each site.

```{r}
#| include: false
library(raster)
library(readr)
library(sf)
library(tidyverse)
library(ggrepel)

raster_url <- "https://github.com/DyerlabTeaching/Raster-Data/raw/main/data/alt_22.tif"
beetle_url <- "https://raw.githubusercontent.com/DyerlabTeaching/Raster-Data/main/data/AraptusDispersalBias.csv"
```

```{r}
#| warning: false
raster <- raster(raster_url)
beetle <- read_csv(beetle_url) %>% 
  st_as_sf(coords = c("Longitude", "Latitude"),
           crs=4326)

bbox <- c(-116, -109, 22, 30)
extent <- extent(bbox)
raster <- crop(raster, extent)

raster %>%
  rasterToPoints() %>%
  as.data.frame() %>% 
  transmute(Longitude = x,
            Latitude = y,
            Elevation = alt_22) %>% 
  ggplot()+ 
  geom_raster(aes(x = Longitude, 
                  y = Latitude,
                  fill = Elevation))+
  geom_sf(data = beetle,
          col = "red")+
  theme_minimal()+
  geom_text_repel(aes(label = Site,
                      geometry = geometry),
                  data = beetle,
                  stat = "sf_coordinates", 
                  size = 4,
                  col = "red")+
  labs(title = "Beetle Observations in Mexico")
```

2. Use the click() function to crop and filter the sites to include only the sampling locations that are in the Cape regions (e.g., south of site labeled #75).  Plot the raster of elevation and all cape region locales with sex-ratio represented as the size of the plot symbol.

```{r}
#| warning: false
plot(raster)
plot(beetle, add = T, col = "red")

point <- click(raster,
               xy = T,
               value = T,
               n = 1) 
point

bbox2 <- c(-116, -109, 22, 24.8)
extent2 <- extent(bbox2)
cropped <- crop(raster, extent2)

beet_crop <- st_crop(beetle,
                     c(xmin = -116, xmax = -109, ymin = 22, ymax = 24.8))

cropped %>% 
  rasterToPoints() %>%
  as.data.frame() %>% 
  transmute(Longitude = x,
            Latitude = y,
            Elevation = alt_22) %>% 
  ggplot()+
  geom_raster(aes(x = Longitude, 
                  y = Latitude,
                  fill = Elevation))+
  scale_fill_gradient(low = "lightblue",
                      high = "orange")+
  geom_sf(data = beet_crop,
          aes(size = MFRatio))+
  labs(title = "Beetles in the Cape Region",
       size = "Males:Females")
```


3. The peninsula of Baja California is divided into the States of Baja California Norte and Baja California Sur. The border between these states is at 28° Latitude. Divide the sample locations into groups based on which state they are located in and plot the average sex ratio of the sample sites partitioned by each site.  

```{r}

```

4. Is there a relationship between the observed sex ratio and the elevation at that site?  Plot these, and in the text, perform a correlation test (cor.test()) and report the values as if you were writing them up in a results section.

```{r}

```

