---
title: "Regression Homework Activity"
author: "Emma Sharp"
format: 
  html:
    code-tools: true
    toc: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library( tidyverse )
library( nycflights13 )

test_boxcox <- function( x, lambdas = seq(-1.1, 1.1, by = 0.015) ) {
  ret <- data.frame( Lambda = lambdas,
                     W = NA,
                     P = NA)
  
  for( lambda in lambdas ) {
    x.tilde <- (x^lambda - 1) / lambda   
    w <- shapiro.test( x.tilde )
    ret$W[ ret$Lambda == lambda ] <- w$statistic
    ret$P[ ret$Lambda == lambda ] <- w$p.value
  }
  
  return( ret )
}
```

Since regression analysis is such a fundamental component of inferences, I thought it would be a good idea to try to work through a few more examples to give you some more practice.  

For this we will use some data on flights into and out of Newark (EWR), JFK, and La Guardia airports in NYC metropolitan area.

```{r}
summary( flights) 
```

We also have some data related to local weather conditions.

```{r}
summary( weather )
```

I'm going to make a synthetic data set consisting of flights out of JFK for the *ExpressJet* airlines.  Since the weather data is on the hour, I also make a synthetic varible `Month.Day.Hour` so that I can join the weather data to this one.

```{r}
flights %>%
  filter( origin == "JFK", carrier == "EV") %>%
  mutate( DateTime = paste(month,day,hour, sep=".")) %>%
  select( DateTime, 
          `Departure Delay` = dep_delay,
          `Departure Time` = sched_dep_time) %>%
  droplevels() -> df.jfk 
summary( df.jfk )
```

Now I'll do the same with the weather data.

```{r}
weather %>%
  filter( origin == "JFK") %>% 
  mutate( DateTime = paste(month,day,hour, sep=".")) %>%
  select( -origin, -year, -month, -day, -hour, -time_hour ) -> df.weather 
summary( df.weather )
```

Now, I'm going to `merge` these two `data.frame` objects using the common `key` I made `DateTime`.

```{r}
merge( df.jfk, df.weather, by="DateTime" ) %>%
  select( -DateTime ) -> df 

summary( df )
```


# Activity

What features of this data set may impact the departure delay for flights coming out of JFK on *ExpressJet* Airlines?

```{r}
#Wind Speed (because the residuals aren't normal, we fail to reject the null hypothesis that there isn't a relationship between wind speed and departure delay)

wspd <- lm(`Departure Delay` ~ wind_speed, data = df)
summary(wspd) #p = 0.01, r^2 = 0.003, F = 6, DF= 1, 1321
plot(wspd, which = 1) #residuals aren't normal

df$wind_speed_adj <- asin(sqrt(df$wind_speed / 100.0))
wspd_adj <- lm(`Departure Delay` ~ wind_speed_adj, data = df)
summary(wspd_adj) #p = 0.04, r^2 = 0.002, F = 4.4, DF = 1, 1321
plot(wspd_adj, which = 1) #residuals still aren't normal

ws <- test_boxcox(df$wind_speed)
df$wspd_lamb <- ((df$wind_speed ^ 0.76) - 1) / 0.76
wspd_lam <- lm(`Departure Delay` ~ wspd_lamb, data = df)
summary(wspd_lam) # p = 0.02, r^2 = 0.003
plot(wspd_lam, which = 1) #residuals still aren't normal

#Wind Gust (not significant, not enough support to reject the null hypothesis that there isn't a relationship between wind gust speed and departure delay)

gust <- lm(`Departure Delay` ~ wind_gust, data = df)
summary(gust) #p = 0.55, r^2 = -0.003, F = 0.36, DF = 1, 238

#Precipitation (residuals aren't normal so we fail to reject the null hypothesis that there is a relationship between precipitation and departure delay)

precip <- lm(`Departure Delay` ~ precip, data = df)
summary(precip) #p << 0, r^2 = 0.01, F = 16.65, DF = 1, 1323
plot(precip, which = 1) #residuals aren't normal 

df$precip_adj <- asin(sqrt(df$precip / 100.0))
precip_adj <- lm(`Departure Delay` ~ precip_adj, data = df)
summary(precip_adj) # p << 0, r^2 = 0.02, F = 30.13
plot(precip_adj, which = 1) #residuals still aren't normal

ps <- test_boxcox(df$precip)
df$precip_lamb <- ((df$precip ^ 0.175) - 1) / 0.175
precip_lam <- lm(`Departure Delay` ~ precip_lamb, data = df)
summary(precip_lam) # p << 0, r^2 = 0.03
plot(precip_lam, which = 1) #residuals not normal

#Pressure (because the residuals aren't normal we fail to reject the null hypothesis)

press <- lm(`Departure Delay` ~ pressure, data = df)
summary(press) #p = 0.001, r^2 = 0.007, F = 9.78, DF = 1, 1207
plot(press, which = 2) #residuals aren't normal 

#Visibility (residuals not normal so we can't reject the null hypothesis)

vis <- lm(`Departure Delay` ~ visib, data = df)
summary(vis) #p << 0, r^2 = 0.03, F = 36.12, DF = 1, 1323
plot(vis, which = 1) #residuals aren't normal

df$vis_adj <- asin(sqrt(df$visib / 100.0))
vis_adj <- lm(`Departure Delay` ~ vis_adj, data = df)
summary(vis_adj) #p << 0, r^2 = 0.3
plot(vis_adj, which = 1) #not normal

vc <- test_boxcox(df$vis_adj)
df$vis_lamb <- ((df$visib ^ 1.090) - 1) / 1.090
vis_lam <- lm(`Departure Delay` ~ vis_lamb, data = df)
summary(vis_lam) # p<<0, r^2 = 0.03
plot(vis_lam, which = 1) #residuals not normal
```



