---
title: "Water Quality in Back Bay National Wildlife Refuge"
author: "Emma Sharp"
date: "09/29/2024"
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

## Abstract

Abstract shit, do later

## Introduction

The data used in this analysis comes from the U.S. Fish and Wildlife Service. The data shows water conditions at Virginia Beach's Back Bay since the 1990s. Back Bay is a National wildlife preserve in the Virginia Beach area. There are five different collection sites; inside Back Bay itself, A-pool, B-pool, C-pool, and D-pool (used for fishing). The data collected at each test site include; turbidity, dissolved oxygen, pH, salinity, and temperature.

## Methods

i love methods

```{r, include=FALSE}
water <- read.csv("BKB_WaterQualityData_2020084.csv",
                stringsAsFactors = T)
require(tidyverse)
library(knitr)
library(kableExtra)
```

```{r}
table <- water %>% 
  mutate(Date = mdy(Read_Date),
         `Site ID` = str_to_title(Site_Id)) %>% 
  filter(Date >= "1980-12-12") %>% 
  select(Date,
         `Site ID`,  
         Salinity..ppt.,
         Water.Depth..m.,
         Water.Temp...C.,
         AirTemp..C.) %>% 
  group_by(`Site ID`) %>% 
  summarise(`Mean Water Temp (C)` = mean(Water.Temp...C.,
                                       na.rm = T),
            `Mean Air Temp (C)` = mean(AirTemp..C.),
            `Mean Salinity (ppt)` = mean(Salinity..ppt.,
                                         na.rm = T)) %>% 
  kable() %>% 
  kable_minimal()
  
```

## Results

```{r, echo=FALSE}
table
```

describe the results (give titleeee)

## Citation

<https://ecos.fws.gov/ServCat/Reference/Profile/117348>
